Exploring the Vectorization of Python Constructs Using Pythran and Boost SIMD

### 2013-11-29 doc/papers/wpmvp14/paper.tex
Update whole version

\title{Vectorizing Python Constructs Using Pythran and Boost SIMD}

@@

The Python language has grown in audience for the past ten years, even reaching
the world of scientific computations thanks to the \texttt{numpy} module, a
%% what about the scipy module? It is not really used?
module that provides a MATLAB-like API. As a consequence, more and more code is
being written either in pure Python, generally to prototype an application, or
as a Python and native code mix when performance matters. However, Python
trades performance for dynamicity and does not particularly shines in terms of
performance.

For instance, the natural solution to Project Euler sixth problem ``Find the
difference between the sum of the squares of the first one hundred natural
numbers and the square of the sum`` can be coded as in
Listing~\ref{lst:euler06-py} in Python and as in Listing~\ref{lst:euler06-c} in
C. A comparison of the performance of the Python function and the C function
called through the \texttt{ctypes} module shows that the C version runs more
than $\times250$ faster than the Python version. Enabling compiler
auto-vectorization makes the C version $\times400$ faster than the Python
version.~\footnote{To perform more reliable time measurements, \texttt{n} has
    been increased to 100001 instead of 101. The script used to perform the
    measures as well as all other experimental data are available at
\url{https://github.com/serge-sans-paille/pythran/tree/wpmvp14/doc/papers/wpmvp14/experiments}.}

@@

Several approaches have been proposed to solve these issues: using native
modules, relying on Ahead-Of-Time (AOT) or Just-In-Time (JIT) compilation.

A native module is a shared library whose interface make it callable from the
Python interpreter using the regular \texttt{import} mechanism. It makes it
possible to call native code from Python in a transparent way. Typically
%% It may be others languages, isn't it?
Python object are unboxed when transferred from Python to C, where efficient
computations can happen. Then the result is boxed when getting back to the
Python world. In between, vectorization can happen. This is the approach taken
by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
contiguous unboxed elements, and provides a great deal of high-level functions
to manipulate them. All the computation-intensive part is done in pure C.
Vectorization can eventually happen at this level, for instance when performing
a point-to-point operation like the sum of two arrays. However the granularity
of the operations limits the ratio of \texttt{load}/\texttt{store} versus operation. For
instance, if \texttt{a}, \texttt{b} and \texttt{c} are 1-D arrays, the sum
\texttt{a + b + c} is computed using two loops and an intermediate array.
% It means there are 4 load / stores and only two loops of addition.

AOT compilation uses type inference or type annotation to statically compute
the type of all (like in Shedskin, Parakeet) or part of (like in Cython or
Numba) variables, effectively unboxing them to speed-up computations. Using JIT
compilation, type information is available at runtime and the unboxing decision
is based on profitability, e.g.\ according to a trace analysis. This is the
approach of the PyPy project. In both cases, generation of vector instruction
is possible, but none of the cited project use it. Indeed automatic
vectorization is a complex topic, and as stated in~\cite{maleki2011}, many cases
are not yet correctly supported even for mainstream compilers.

Pythran is also an AOT compiler, which holds the unique property of turning
Python modules into C++ \emph{meta-programs}. It does not need type annotations
and keeps the polymorphism of original functions, respecting Python's duck
typing. Its compilation workflow is described in Figure~\ref{fig:pythran-compiler}.
Like others AOT compilers, it unboxes all variables for better performance, and
performs a wide variety of optimizations such as lazy evaluation, constant
folding, expression fusion etc. It supports core \texttt{numpy} constructs and
most Python constructs to the notable exception of user classes. Constructs
such as \texttt{eval}, are not supported.

@@

This paper introduces vectorization of Python programs based on the idea that
many Python constructs already exhibit good vectorization opportunities. The
goal of the paper switches from automatic vectorization to vectorization of
high-level Python intrisics, as described in Section~\ref{sec:python-semantic}.
One of the difficulties of vectorizing Python functions is linked to function
polymorphism. Vectorization in the context of C++ meta-programs is examined in
Section~\ref{sec:meta-vectorization}. The approach is validated using the
Pythran compiler on several micro-applications from signal processing, physics
and mathematics presented in Section~\ref{sec:benchs}.

%%
%%
\section{Vectorization of a Meta-Program}
\label{sec:meta-vectorization}
SIMD extensions have been a feature of choice for processor manufacturers for a
couple of decades. However, programming applications that take advantage of the 
SIMD extension available on the current target remains a complex task. Programmers 
that use low-level intrinsics have to deal with a verbose programming style due 
to the fact that SIMD instructions sets cover a few common functionnalities, requiring
to bury the initial algorithms in architecture specific implementation details.
Furthermore, these efforts have to be repeated for every different extension
that one may want to support, making design and maintenance of such applications 
very time consuming.

@@

\subsection{\texttt{Load}/\texttt{Store} Protocol}

@@

A high-level language such as Python exhibit another level of vectorization
opportunity in implict loop. Implicit loops can be found in many places, but
this paper focuses on \emph{list comprehension} and \emph{generator
expression}, \emph{map} or \emph{sum} intrinsics and \emph{array operations}
through the \texttt{numpy} module.

@@

\subsubsection{\texttt{sum}}
% only sum as texttt in a title is not really beautiful

The \texttt{sum} intrinsic is a typical case of potential vectorization of an
implict loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
value} and its behavior is equivalent to the Python code in
Listing~\ref{lst:sum-py}. Depending on the type of \texttt{sequence}, and also
wether one needs to strictly conforms to the IEEE Standard for Floating-Point
Arithmetic (IEEE 754).
% I don't understand what this last sentence is doing here :-s

@@

As function have to be generic but some containers do not match requierement
properties, it is possible to select a vectorized
implementation of fallback to a generic one using Substitution Failure
Is Not An Error (SFINAE).

@@

To validate the approach, a number of code snippets presenting vectorization
opportunities have been selected and benchmarked. The original Python version
is used as a reference, then code is compiled with Pythran without
vectorization support and finally with Pythran and vectorization support. Time
measurements are made using the \texttt{timeit} module.

@@

@@

To evaluate the vectorization of the \texttt{map} intrisinc, two expressions
are used: \texttt{map(abs, r)} and \texttt{map(lambda x: 5 * x + 3, r)}, where
\texttt{r} is a \texttt{ndarray} of random double-precision floating point
numbers. Table~\ref{tbl:map-timings} summarizes the result. The execution time
for the Cython version is extremly long due to the ndarray indexing and
function call overheads in CPython. Thanks to static typing and aggressive
inlining, Pythran does not suffer from these issues, as shown by its sequential
execution time. Vectorization is beneficial and yields a $\times2.93$ speedup.

\begin{table}

    \centering
    \begin{tabular}{|l|r|r|r|}
        \hline
                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
        \hline
        \texttt{map(abs, r)}                    &   1530ms  & 1.72ms    & 0.587ms   \\
        \hline
        \texttt{map(f, r)}    &   141ms   & 1.72ms    & 0.587ms  \\
        \hline

    \end{tabular}
    \caption{Timings of various \texttt{map} calls using CPython, Pythran, and Pythran with vectorization.}
    \label{tbl:map-timings}

\end{table}

@@

To evaluate the vectorization of the \texttt{sum} intrisinc, two expressions
are used: \texttt{sum(r)} and \texttt{sum(r * r)}, where \texttt{r} is a
\texttt{ndarray} of random double-precision floating point numbers.
Table~\ref{tbl:sum-timings} summarizes the result. Again, CPython is extremely
slow. It is interesting to note that the extra array multiplication does not
impacts the excution time much, simply because it is an operation pefromed at
the native level by the \texttt{numpy} module, while the \texttt{sum} is fully
interpreted. The vectorized reductions yields a decent $\times2.41$ speedup
over the sequential version.

\begin{table}

    \centering
    \begin{tabular}{|l|r|r|r|}
        \hline
                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
        \hline
        \texttt{sum(r)}                    &   277ms  & 1.07ms    & 0.416ms   \\
        \hline
        \texttt{sum(r * r)}    &   292ms   & 1.09ms    & 0.452ms  \\
        \hline

    \end{tabular}
    \caption{Timings of various \texttt{sum} calls using CPython, Pythran, and Pythran with vectorization.}
    \label{tbl:sum-timings}

\end{table}

\subsection{Vectorization of \texttt{ndarray} point-to-point operations}

To evaluate the vectorization of \texttt{ndarray} point-to-point operations,
three expressions are used: \texttt{1. / a + a}, the example from
Listing~\ref{lst:numpy-code} and a computation of the point-to-point arc
distance  between random values. Again, all inputs are large randomized
\texttt{ndarray} of double precision floating point numbers.
Table~\ref{tbl:numpy-timings} summarizes the results. Because all computations
are done in native space, there is not much speedup when running them through
Pythran. The only benefits comes from the elimination of temporary arrays and
loop fusion. As a consequence, the more elements in the \texttt{numpy}
expression, the more Pythran accelerates the code. The vectorization speedups
follow the same process, as larger expression tend to increase the operation
over memory movement ratio.

\begin{table}

    \centering
    \begin{tabular}{|l|r|r|r|}
        \hline
                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
        \hline
        \texttt{1. / a + a}                    &   7.48ms  & 6.19ms    & 4.05ms   \\
        \hline
        \emph{rosen}                           &   13.9ms   & 2.98ms   & 1.15ms  \\
        \hline
        \emph{arc distance}                    &   178ms   & 146ms    & 53.8ms  \\
        \hline

    \end{tabular}
    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
    \label{tbl:numpy-timings}

\end{table}

This experiment considers the code from Listing~\ref{lst:euler06-py}.

\begin{table}

    \centering
    \begin{tabular}{|l|r|r|r|}
        \hline
                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
        \hline
        \emph{euler06}                    &   12ms   & 0.34ms    & 0.30ms  \\
        \hline

    \end{tabular}
    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
    \label{tbl:numpy-timings}

\end{table}

@@

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.
\cite{*}

>>>>

\title{Exploring the Vectorization of Python Constructs Using Pythran and Boost SIMD}

@@

The Python language~\cite{rossum97} has grown in audience for the past ten
years, even reaching the world of scientific computations~\cite{oliphant2007}
thanks to the \texttt{numpy} module~\cite{numpyarray2011}, a module that
provides an efficient multi-dimensionnal array type, and the \texttt{scipy}
package~\cite{scipy} that provides a MATLAB-like API. As a consequence, more
and more code is being written either in pure Python, generally to prototype an
application, or as a Python and native code mix when performance matters.
However, Python trades performance for dynamicity and does not particularly
shines in terms of performance.

\subsection{Python Performance Shortcomings}

For instance, the natural solution to Project Euler sixth problem ``Find the
difference between the sum of the squares of the first one hundred natural
numbers and the square of the sum`` can be coded as in
Listing~\ref{lst:euler06-py} in Python and as in Listing~\ref{lst:euler06-c} in
C. A comparison of the performance of the Python function and the C function
called through the \texttt{ctypes} module shows that the C version runs more
than $\times250$ faster than the Python version. Enabling compiler
auto-vectorization makes the C version $\times400$ faster than the Python
version. To perform more reliable time measurements, \texttt{n} has been
increased to 100001 instead of 101.

@@

\subsection{Compiling Python Code}

Several approaches have been proposed to solve the performance issues of
Python: using native modules, relying on Ahead-Of-Time (AOT) or Just-In-Time
(JIT) compilation.

A native module is a shared library whose interface make it callable from the
Python interpreter using the regular \texttt{import} mechanism. It makes it
possible to call native code from Python in a transparent way. Typically Python
object are unboxed when transferred from Python to C, where efficient
computations can happen. Then the result is boxed when getting back to the
Python world. In between, vectorization can happen. This is the approach taken
by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
contiguous unboxed elements, and provides a great deal of high-level functions
to manipulate them. All the computation-intensive part is done in pure C.
Vectorization can eventually happen at this level, for instance when performing
a point-to-point operation like the sum of two arrays. However the granularity
of the operations limits the ratio of \texttt{load}/\texttt{store} versus
operation. For instance, if \texttt{a}, \texttt{b} and \texttt{c} are 1-D
arrays, the sum \texttt{a + b + c} is computed using two loops and an
intermediate array.  It means there are 4 load / stores and only two additions
per loop iteration.

AOT compilation uses type inference or type annotation to statically compute
the type of all (like in Shedskin~\cite{shedskin2006},
Parakeet~\cite{parakeet2012}) or part of (like in Cython~\cite{cython2010} or
Numba~\cite{numba}) variables, effectively unboxing them to speed-up
computations. Using JIT compilation, type information is available at runtime
and the unboxing decision is based on profitability, e.g.\ according to a trace
analysis. This is the approach of the PyPy project~\cite{pypy2009}. In both
cases, generation of vector instruction is possible, but none of the cited
project use it. Indeed automatic vectorization is a complex topic, and as
stated in~\cite{maleki2011}, many cases are not yet correctly supported even
for mainstream compilers.

\subsection{The Pythran Compiler}

Pythran~\cite{pythran2013, pyhpc2013} is also an AOT compiler, bit it holds the
unique property of turning Python modules into C++ \emph{meta-programs}. It
does not need type annotations and keeps the polymorphism of original
functions, respecting Python's duck typing. Its compilation workflow is
described in Figure~\ref{fig:pythran-compiler}.  Like others AOT compilers, it
unboxes all variables for better performance, and performs a wide variety of
optimizations such as lazy evaluation, constant folding, expression fusion etc.
It supports core \texttt{numpy} constructs and most Python constructs to the
notable exception of user classes. Constructs such as \texttt{eval}, are not
supported either.

@@

This paper explores the vectorization of Python programs based on the idea that
many Python constructs already exhibit good vectorization opportunities. The
goal of the paper switches from automatic vectorization to vectorization of
high-level Python intrinsics, as described in Section~\ref{sec:python-semantic}.
One of the difficulties of vectorizing Python functions is linked to function
polymorphism. Vectorization in the context of C++ meta-programs is examined in
Section~\ref{sec:meta-vectorization}. The approach is validated using the
Pythran compiler on several kernels presented in Section~\ref{sec:benchs}.

%%
%%
\section{Vectorization of a Meta-Program}
\label{sec:meta-vectorization}

Single Instruction Multiple Data (SIMD) extensions have been a feature of
choice for processor manufacturers for a couple of decades. However,
programming applications that take advantage of the SIMD extension available on
the current target remains a complex task. Programmers that use low-level
intrinsics have to deal with a verbose programming style due to the fact that
SIMD instructions sets cover a few common functionnalities, requiring to bury
the initial algorithms in architecture specific implementation details.
Furthermore, these efforts have to be repeated for every different extension
that one may want to support, making design and maintenance of such
applications very time consuming.

@@

\subsection{Interaction with Pythran}

The first attempt to have Pytrhan generate vectorized code directly relied on
SSE intrinsics. This approach quickly proved to be inappropriate for several
reasons:

    \item[portability] Only SSE was supported, moving to AVX or another vector
        instruction set would require a great deal of engineering;

    \item[completness] Many basic mathematic functions such as cosinus or sinus
        are not directly available in the target instruction sets;

    \item[typing] Pytrhan manipulates meta-functions with parametric types,
        while a given instruction set only provide concretes type.

The \boostsimd{} abstraction layer solves these three issues to the expense of
an increase in compilation time, which still remains acceptable for an AOT
compiler.

Pythran uses function objects to represent any Python functions, either
user-functions or functions from the standard library. The former are
automatically generated from the user code analysis, and the latter are
hand-coded in a library called \texttt{pythonic++} provided with the compiler.
For instance the Python function \texttt{def foo(x): return 3 * x} is compiled
into a code similar to the one in Listing~\ref{lst:simple-pythran-output}. The
actual would be more compilcated to handle type inference across several
statements, but the structure remains valid.

Thanks to this representation, both the Python and C++ function can be called
on scalar types, \texttt{list} type or any other type that supports
multiplication by an integer constant. The only difference is that the Python
version uses \emph{dynamic polymorphism} while the C++ version uses
\emph{static polymorphism}.

\begin{lstlisting}[language=c++, caption={Simplified Pythran Translation of a Polymorphic Function}, label={lst:simple-pythran-output}]
struct foo {
    template<class T>
    auto operator()(T&& x)
    -> decltype(3 * x)
    {
        return x * 3;
    }
};
\end{lstlisting}

This representation perfectly matches the requirements of \boostsimd, as
presented in Listing~\ref{lst:square-cxx}. The \texttt{foo} function can be
used indifferently on scalar types and vector types.

@@

A high-level language such as Python exhibits another level of vectorization
opportunity in implict loop. Implicit loops can be found in many places, but
this paper focuses on \emph{list comprehension} and \emph{generator
expression}, \emph{map} or \emph{sum} intrinsics and \emph{array operations}
through the \texttt{numpy} module.

@@

\subsubsection{The \texttt{sum} Intrinsic}

The \texttt{sum} intrinsic is a typical case of potential vectorization of an
implict loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
value} and its behavior is equivalent to the Python code in
Listing~\ref{lst:sum-py}. Depending on the type of \texttt{sequence}, and also
wether one needs to strictly conforms to the IEEE Standard for Floating-Point
Arithmetic (IEEE 754), vectorization of this reduction is a comm case of \texttt{fold}.

@@

As function have to be generic but some containers do not match requierement
properties, it is possible to select a vectorized
implementation of fallback to a generic one using Substitution Failure
Is Not An Error (SFINAE)~\cite{metaprogramming2002}.

@@

To validate the approach, a number of code snippets presenting vectorization
opportunities have been selected and benchmarked. The original Python version
is used as a reference, then code is compiled with Pythran without
vectorization support and finally with Pythran and vectorization support. Time
measurements are made using the \texttt{timeit} module, which runs the piece of
code a number of time and measure its average execution time. It does this
three times and picks the best average execution time among those.

@@

The scripts used to perform the measures as well as all other experimental data
used in this paper are available online at
\url{https://github.com/serge-sans-paille/pythran/tree/wpmvp14/doc/papers/wpmvp14/experiments}.

This experiment considers the code from Listing~\ref{lst:euler06-py}. This code
uses \emph{generator expression} that have not been explored in this paper, so
it has been modified to use \emph{list comprehension} instead. As a result, a
lot of temporay array are used, which leads to the performance show on
Figure~\ref{fig:euler06-timings}. Although Pythran does a decent job at
optimizing the code, vectorization does not provide significant speedup. This
is due to the poor computation over memory transfer ratio, and the lack of
optimization done by Pythran on that aspect for \texttt{list}s. A similar setup
is evaluated in the context of \texttt{numpy} arrays in
\S~\ref{sec:bench-numpy} that use temporary array elimination.

\begin{figure}[ht]

    \includegraphics[width=.5\textwidth]{euler06}
    \caption{Timings of euler06 problem using CPython, Pythran, and Pythran with vectorization.}
    \label{fig:euler06-timings}

\end{figure}

@@

To evaluate the vectorization of the \texttt{map} intrisinc, two expressions
are used: \texttt{map(abs, r)} and \texttt{map(lambda x: 5 * x + 3, r)}, where
\texttt{r} is a \texttt{ndarray} of random double-precision floating point
numbers. Figure~\ref{fig:map-timings} summarizes the result. The execution time
for the Cython version is extremly long due to the ndarray indexing and
function call overheads in CPython. Thanks to static typing and aggressive
inlining, Pythran does not suffer from these issues, as shown by its sequential
execution time. Vectorization is beneficial and yields a $\times2.93$ speedup.

\begin{figure}[ht]

    \includegraphics[width=.5\textwidth]{map_abs}
    \includegraphics[width=.5\textwidth]{map_fun}
    \caption{Timings of various \texttt{map} calls using CPython, Pythran, and Pythran with vectorization.}
    \label{fig:map-timings}

\end{figure}

@@

To evaluate the vectorization of the \texttt{sum} intrisinc, two expressions
are used: \texttt{sum(r)} and \texttt{sum(r * r)}, where \texttt{r} is a
\texttt{ndarray} of random double-precision floating point numbers.
Figure~\ref{fig:sum-timings} summarizes the result. Again, CPython is extremely
slow. It is interesting to note that the extra array multiplication does not
impacts the excution time much, simply because it is an operation pefromed at
the native level by the \texttt{numpy} module, while the \texttt{sum} is fully
interpreted. The vectorized reductions yields a decent $\times2.41$ speedup
over the sequential version.

\begin{figure}[ht]

    \includegraphics[width=.5\textwidth]{sum}
    \includegraphics[width=.5\textwidth]{sum_square}
    \caption{Timings of various \texttt{sum} calls using CPython, Pythran, and Pythran with vectorization.}
    \label{fig:sum-timings}

\end{figure}

\subsection{Vectorization of \texttt{ndarray} point-to-point operations}
\label{sec:bench-numpy}

To evaluate the vectorization of \texttt{ndarray} point-to-point operations,
three expressions are used: \texttt{1. / a + a}, the example from
Listing~\ref{lst:numpy-code} and a computation of the point-to-point arc
distance  between random values. Again, all inputs are large randomized
\texttt{ndarray} of double precision floating point numbers.
Figure~\ref{fig:numpy-timings} summarizes the results. Because all computations
are done in native space, there is not much speedup when running them through
Pythran. The only benefits comes from the elimination of temporary arrays and
loop fusion. As a consequence, the more elements in the \texttt{numpy}
expression, the more Pythran accelerates the code. The vectorization speedups
follow the same process, as larger expression tend to increase the operation
over memory movement ratio.

\begin{figure}[ht]

    \includegraphics[width=.5\textwidth]{numexpr}
    \includegraphics[width=.5\textwidth]{rosen}
    \includegraphics[width=.5\textwidth]{arc}
    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
    \label{fig:numpy-timings}

\end{figure}

\section*{Conclusion}

This paper explores the design and implementation of a vectorizer for the
Python language. It targets high-level constructs from the language and from
the \texttt{numpy} module that already have a vector semantic, efficiently
vectorizing them thanks to the combination of Pythran, a Python to C++ compiler
that turns Python modules into C++ meta-programs and \boostsimd, a C++ library
to program SIMD architectures.

Common Python constructions such as \emph{list comprehension}, the \texttt{map}
and \texttt{sum} intrinsics and array operations from \texttt{numpy} are
considered and the approach is validated on several kernels, showing
promising results.

Future works include the case of Python generators, that have the benfit of not
creating intermediate container to hold the data, but require more analyse in
the context of vectorization. Handling \texttt{list} operations in an efficient
manner as done for \texttt{numpy} arrays is also a required developpment axis.

@@

Several students from Télécom Bretagne have been working on this project, the
authors would like to thank Alan Raynaud, Adrien Merlini and Yuanchang Peng for
their contribution to the Pytrhan project. They are also in debt to Mehdi
Aminin for his careful code reviews.

% We recommend abbrvnat bibliography style.
